Transition path sampling based calculations of free energies of enzymatic reactions: the case of human methionine adenosyl transferase and plasmodium vivax adenosine deaminase

### 2022-04-29 tex/manuscript.tex
added more text and figures

Frenkel has shown that inclusion of rejected states in Markov state Monte 
Carlo algorith can help speed up the convergence of the calculations of 
order-parameter distribution of a two-dimensional
Ising model. \cite{Frenkel04ProcNatAcadSci101p17571}
Inclusion of rejected trajectories within TPS was first discussed by Radhakrishnan 
et al. \cite{Radhakrishnan04JChemPhys121p2436} who developed `BOLAS' which combines 
the shooting and shifting algorithms of TPS with a window based sampling technique 
in the spirit of umbrella sampling to calculate equilibrium free energy 
differences. Peters et al. used a variant of the BOLAS scheme called the 
equilibrium path sampling (EPS) where they use aimless shooting 
combined with window based sampling for free energy calculations. 
\cite{Peters08JAmChemSoc130p17342,Beckham10epsbook} 
More recently Brotzakis et al., \cite{Brotzakis19JChemPhys151p174111}
have developed a reweighted path ensemble scheme within the standard TPS method 
called virtual interface exchange TPS (VIE-TPS) to calculate the free energy 
landscape. VIE-TPS is shown to produce reasonable free energies for 
diffusive reactions whereas it is less accurate otherwise.

@@

%Monte carlo methods in trajectory space the probability of a path is 
%\begin{equation}
%P(x_A)
%\end{equation}

@@

et al. \cite{Niland21Biochem60p791} and the crystal structure of \textit{pv}ADA 
in complex with MT-coformycin (protein data bank ID: 3EWC) reported by Ho et 
al. \cite{Ho09Biochemistry48p9618} as the starting points of our two 

@@

The energies of both the systems were minimized using 50 steps of the 
steepest descent method, followed by 2000 steps of the
adopted basis Newton-Raphson method where only classical molecular mechanics 
was used for the dynamics. 
The minimized systems were
heated slowly to 300 K for 35 ps beginning with harmonic
constraints on all atoms except on the H atoms and the TIP3P 
water molecules with gradual reduction of the restraint forces to 1. 
15 ps of equilibration was carried out starting with harmonic 
constraints of force constant 1 followed by 20 ps of contraint free 
equilibration to prepare the systems for TPS simulations. 
During the heating and equilibration steps, bonds that contain 
hydrogen atoms in the MM region were restrained to their equilibrium values 
using the SHAKE procedure. \cite{Ryckaert77JComputPhys23p327}

%------------------------------------------------------------------------------ 
\subsection{Transition path sampling}
%------------------------------------------------------------------------------
TPS simulations require the definition of unique reactant and product states
to collect reactive trajectories based on the Monte Carlo algorithm. Here we 
discuss the details of the reactions involving the MAT2A and the $pv$ADA enzymes 
to identify and define the order parameters characterizing the reactant and product basins.

@@

A biased initial reactive trajectory is obtained by using harmonic restraint forces 
of 1500 to break the $O-C$ bond and form the $S-C$ bond for the MAT2A enzyme and 
to break the $N-C$ bond and form the $O-C$ bond for the case of the $pv$ADA enzyme. 
Using the biased trajectory as the starting point and after removing the restraint 
forces, the shooting algorithm was used to generate
new trajectories within which first a random time slice is chosen from the reactive 
trajectory and perturbing the momenta of the system. 
The time step that we used for all our calculations was $1\;fs$ and we propagate the 
dynamics for a total of 250 steps forward and 250 steps backward which amounts to a 
total of $0.5\;ps$. 
A total of 150 and 100 reactive trajectories were generated for the MAT2A and $pv$ADA
enzymes, respectively.

%------------------------------------------------------------------------------ 
\subsection{Free energy calculations}
%------------------------------------------------------------------------------ 
To obtain the free energy profile along the order parameter, we use the method 
discussed in sec. \ref{ssec:algorithm}. The order paratemer for the MAT2A system 
is defined to be $dis(OC)-dis(SC)$ and that for the $pv$ADA is defined as 
$dis(NC)-dis(OC)$. For the MAT2A system, configurations are sampled within 
the range $[-2.0,2.0]$ and this range is divided into 14 overlapping windows 
($0.09$ {\AA} overlap on either side). 
Configurations are sampled for the $pv$ADA system within the interval $[-3.0, 3.0]$ {\AA}
of the order parameter value. 
This interval is divided into 20 overlapping windows ($0.045$ {\AA} between 
neighboring windows). A reactive trajectory from the TPS ensemble is chosen to be the guide trajectory 
for sampling within each of the windows. To sample in a particular window $w$, a 
frame from this reactive trajectory is chosen such that the value of its 
order parameter falls within $w$. Using the shooting algorithm
the momenta are perturbed and dynamics is run for 10 steps forwards and backwards with a 
$1\;fs$ time step to obtain a trajectory with a time period of $20\;fs$.    
Within each window 3000 TPS trajectories are collected. The 
probability distribution function for the order parameter within window $w$ 
is then calculated as normalized histograms which represent the populations 
of the configurations within the window at equilibrium. 
Then we calculate the free energy profile in a similar fashion to the WHAM 
procedure \cite{Kumar92JComputChem13p1011} which is generally used to obtain the 
potential of mean force from umbrella sampling calculations. 
%------------------------------------------------------------------------------ 
\section{Results and discussion}
%------------------------------------------------------------------------------ 
\subsection{Human Methionine Adenosyltransferase 2A}
\begin{figure}[ht!]
\includegraphics[scale=0.2]{figures/mat2a-equil.png}
\label{fig:mat2a-equil}
\caption{Equilibriated structure of MAT2A enzyme}
\end{figure}

The binding pocket image is shown for this enzyme here
% width: 3.330, height: 1.873
The TPS ensemble was used to analyze the reaction pathway for the S$_{\text{N}}2$
mechanism and a typical trajectory is provided here
\begin{figure}
\includegraphics[scale=1.0]{figures/mat2a-diff60.pdf}
\caption{A reactive trajectory}
\label{fig:mat2a-reactive-traj}
\end{figure}
The order parameters of the system along a typical reactive trajectory is 
shown in Fig. \ref{fig:mat2a-reactive-traj}.

@@

\begin{figure}[ht!]
\includegraphics[scale=1.0]{figures/comm-60-mat2a.pdf}
\end{figure}

\begin{figure}
\includegraphics[scale=1.0]{figures/mat2a-fenergy.pdf}
\caption{The free energy as a function of the order parameter.}
\end{figure}

%------------------------------------------------------------------------------ 
\subsection{Adenosine deaminase reaction coordinates and free energy}
%------------------------------------------------------------------------------ 
\begin{figure}[ht!]
\centering
\includegraphics[scale=0.2]{./figures/ada-equil.png}
\caption{Equilibriated structure of ADA protein}
\end{figure}

\begin{figure}[ht!]
\includegraphics[scale=1.0]{./figures/ada-fenergy.pdf}
\label{fig:ada-fenergy}
\caption{The free energy curve for the ADA reaction}
\end{figure}

%\begin{figure}[ht]
%\centering
%\begin{minipage}[b]{0.32\linewidth}
%\includegraphics[scale=0.3]{figures/reactant-state.png}
%\caption{Reactant state}
%\label{fig:minipage1}
%\end{minipage}
%\quad
%\begin{minipage}[b]{0.32\linewidth}
%\includegraphics[scale=0.3]{figures/int1.png}
%\caption{Intermediate state}
%\label{fig:minipage2}
%\end{minipage}
%\quad
%\begin{minipage}[b]{0.32\linewidth}
%\includegraphics[scale=0.3]{figures/product-state.png}
%\caption{Product state}
%\label{fig:minipage1}
%\end{minipage}
%\end{figure}
%------------------------------------------------------------------------------ 
\section{Conclusions}
%------------------------------------------------------------------------------ 
We have discussed the application of a scheme to calculate the 
free energies of enzymatic reactions within the TPS method using the MAT2A
and the ADA enzymes as examples.

>>>>

Frenkel has shown that inclusion of rejected states in Markov state Monte 
Carlo algorith can help speed up the convergence of the calculations of 
order-parameter distribution of a two-dimensional
Ising model. \cite{Frenkel04ProcNatAcadSci101p17571}
Inclusion of rejected trajectories within TPS was first discussed by Radhakrishnan 
et al. \cite{Radhakrishnan04JChemPhys121p2436} who developed `BOLAS' which combines 
the shooting and shifting algorithms of TPS with a window based sampling technique 
in the spirit of umbrella sampling to calculate equilibrium free energy 
differences. Peters et al. used a variant of the BOLAS scheme called the 
equilibrium path sampling (EPS) where they use aimless shooting 
combined with window based sampling for free energy calculations. 
\cite{Peters08JAmChemSoc130p17342,Beckham10epsbook} 
More recently Brotzakis et al., \cite{Brotzakis19JChemPhys151p174111}
have developed a reweighted path ensemble scheme within the standard TPS method 
called virtual interface exchange TPS (VIE-TPS) to calculate the free energy 
landscape. VIE-TPS is shown to produce reasonable free energies for 
diffusive reactions whereas it is less accurate otherwise.
Transition interface sampling (TIS) has been used to obtain 
reweighted path ensemble which can give access to free energy 
and committor surfaces. \cite{Rogal10JChemPhys17p174109}

@@

@@

et al. \cite{Niland21Biochem60p791,Ghosh21JAmChemSoc143p18325} and the crystal 
structure of \textit{pv}ADA in complex with MT-coformycin (protein data bank ID: 3EWC) 
reported by Ho et al. \cite{Ho09Biochemistry48p9618} as the starting points of our two 

@@

The energies of both the systems were minimized using 50 steps of the 
steepest descent method, followed by 2000 steps of the
adopted basis Newton-Raphson method where only classical molecular mechanics 
was used for the dynamics. 
The minimized systems were
heated slowly to 300 K for 35 ps beginning with harmonic
constraints on all atoms except on the H atoms and the TIP3P 
water molecules with gradual reduction of the restraint forces to 1. 
15 ps of equilibration was carried out starting with harmonic 
constraints of force constant 1 followed by 20 ps of contraint free 
equilibration to prepare the systems for TPS simulations. 
During the heating and equilibration steps, bonds that contain 
hydrogen atoms in the MM region were restrained to their equilibrium values 
using the SHAKE procedure. \cite{Ryckaert77JComputPhys23p327} 
%------------------------------------------------------------------------------ 
\subsection{Transition path sampling}
%------------------------------------------------------------------------------
TPS simulations require the definition of unique reactant and product states
to collect reactive trajectories based on the Monte Carlo algorithm. Here we 
discuss the details of the reactions involving the MAT2A and the $pv$ADA enzymes 
to identify and define the order parameters characterizing the reactant and product basins.

@@

A biased initial reactive trajectory is obtained by using harmonic restraint forces 
of 1500 to break the $O-C$ bond and form the $S-C$ bond for the MAT2A enzyme and 
to break the $N-C$ bond and form the $O-C$ bond for the case of the $pv$ADA enzyme. 
Using the biased trajectory as the starting point and after removing the restraint 
forces, the shooting algorithm was used to generate
new trajectories within which first a random time slice is chosen from the reactive 
trajectory and perturbing the momenta of the system. 
The time step that we used for all our calculations was $1\;fs$ and we propagate the 
dynamics for a total of 250 steps forward and 250 steps backward which amounts to a 
total of $0.5\;ps$. 
A total of 150 and 100 reactive trajectories were generated for the MAT2A and $pv$ADA
enzymes, respectively. 
%------------------------------------------------------------------------------ 
\subsection{Free energy calculations}
%------------------------------------------------------------------------------ 
To obtain the free energy profile along the order parameter, we use the method 
discussed in sec. \ref{ssec:algorithm}. The order paratemer for the MAT2A system 
is defined to be $dis(OC)-dis(SC)$ and that for the $pv$ADA is defined as 
$dis(NC)-dis(OC)$. For the MAT2A system, configurations are sampled within 
the range $[-2.0,2.0]$ and this range is divided into 14 overlapping windows 
($0.09$ {\AA} overlap on either side). 
Configurations are sampled for the $pv$ADA system within the interval $[-3.0, 3.0]$ {\AA}
of the order parameter value. 
This interval is divided into 20 overlapping windows ($0.045$ {\AA} between 
neighboring windows). A reactive trajectory from the TPS ensemble is chosen to be the guide trajectory 
for sampling within each of the windows. To sample in a particular window $w$, a 
frame from this reactive trajectory is chosen such that the value of its 
order parameter falls within $w$. Using the shooting algorithm
the momenta are perturbed and dynamics is run for 10 steps forwards and backwards with a 
$1\;fs$ time step to obtain a trajectory with a time period of $20\;fs$.    
Within each window 3000 TPS trajectories are collected. The 
probability distribution function for the order parameter within window $w$ 
is then calculated as normalized histograms which represent the populations 
of the configurations within the window at equilibrium. 
Then we calculate the free energy profile in a similar fashion to the WHAM 
procedure \cite{Kumar92JComputChem13p1011} which is generally used to obtain the 
potential of mean force from umbrella sampling calculations. 
%------------------------------------------------------------------------------ 
\section{Results and discussion}
%------------------------------------------------------------------------------ 
\subsection{Human Methionine Adenosyltransferase 2A}
The equilibrated structure of the MAT2A enzyme in complex with 
\begin{figure}[ht!]
\includegraphics[scale=0.2]{figures/mat2a-equil.png}
\label{fig:mat2a-equil}
\caption{Equilibriated structure of the MAT2A enzyme in complex with ATP and 
MET along with 2 Mg$^{2+}$ ions.}
\end{figure}
The adenosyl part of the ATP is held together through $\pi-\pi$ interactions 
by Phe250 and Ile117 residues.

\begin{figure}
\includegraphics[scale=1.0]{figures/mat2a-diff167.pdf}
\caption{A reactive trajectory}
\label{fig:mat2a-reactive-traj}
\end{figure}
The order parameters of the system along a typical reactive trajectory is 
shown in Fig. \ref{fig:mat2a-reactive-traj}.

@@

\begin{figure*}[ht!]
\centering
\begin{minipage}[b]{0.45\linewidth}
\includegraphics[width=\textwidth]{figures/comm-60-mat2a-nocons.pdf}
%\caption{}
\label{fig:minipage1}
\end{minipage}
\quad
\begin{minipage}[b]{0.45\linewidth}
\includegraphics[width=\textwidth]{figures/comm-60-mat2a.pdf}
%\caption{}
\label{fig:minipage2}
\end{minipage}
\caption{Committor distribution analysis}
\end{figure*}

The activation energy of this reaction was experimentally measured by 
Niland et al. \cite{Niland21Biochem60p791} where they measured the rate constant of the reaction 
at various temperatures while keeping the substrate concentrations near saturation.
Using the Arrhenius equation they estimated the activation energy to be 
$17.27\;\pm\;1.5\;\text{Kcal mol}^{-1}$. Compared to typical enzyme catalyzed 
reactions calculated activation energy is small.

\begin{figure}
\includegraphics[scale=1.0]{figures/mat2a-fenergy.pdf}
\caption{The free energy as a function of the order parameter.}
\end{figure}
From our TPS based calculations we calculated the free energy barrier to be 
$16\;\text{Kcal mol}^{-1}$.
%------------------------------------------------------------------------------ 
\subsection{Adenosine deaminase reaction coordinates and free energy}
%------------------------------------------------------------------------------ 
The equilibrated structure for the ADA enzyme complexed with adenosine, 
OH$^{-}$ anion and Zn$^{2+}$ ion is shown in Fig. \ref{fig:ada-equil}.
The Zn$^{2+}$ cation interacts with His42, His44 and His226 residues
along with the hydroxide ion in the pocket. Glu229 is part of the 
QM region and it acts as a proton donor to the adenosine molecule.

\begin{figure}[ht!]
\centering
\includegraphics[scale=0.2]{./figures/ada-equil.png}
\caption{Equilibriated structure of ADA protein}
\label{fig:ada-equil}
\end{figure}

\begin{figure}
\includegraphics[scale=1.0]{figures/ada-diff60.pdf}
\caption{A reactive trajectory}
\label{fig:ada-reactive-traj}
\end{figure}

\begin{figure}[ht!]
\includegraphics[scale=1.0]{./figures/ada-fenergy.pdf}
\label{fig:ada-fenergy}
\caption{The free energy curve for the ADA reaction}
\end{figure}
%------------------------------------------------------------------------------ 
\section{Conclusions}
%------------------------------------------------------------------------------ 
We have discussed the application of a scheme to calculate the 
free energies of enzymatic reactions within the TPS method using the MAT2A
and the ADA enzymes as examples.
